---
title: "Trustworthy and Adaptable Data Integration"
author: "Kris Sankaran"
output:
  xaringan::moon_reader:
    css: ["default", "css/xaringan-themer.css"]
    lib_dir: libs
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
    seal: false
---

class: title

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(MASS)
library(knitr)
library(RefManageR)
library(tidyverse)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, dpi = 200, fig.align = "center", fig.width = 6, fig.height = 3)
min_theme <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    strip.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(min_theme)

# overwrite some default scales in ggplot2
scale_fill_continuous <- function(...) scico::scale_fill_scico(..., palette = "lapaz", direction = -1)
scale_colour_discrete <- function(...) ggplot2::scale_color_brewer(..., palette = "Set2")
scale_x_continuous <- function(...) ggplot2::scale_x_continuous(..., expand = c(0, 0))
scale_y_continuous <- function(...) ggplot2::scale_y_continuous(..., expand = c(0, 0))

BibOptions(
  check.entries = TRUE, 
  bib.style = "authoryear", 
  cite.style = "authoryear", 
  style = "markdown",
  hyperlink = FALSE, 
  dashed = FALSE,
  max.names = 1
)
bib <- ReadBib("references.bib")
```

<div id="title">
Trustworthy and Adaptable Data Integration
</div>
<div id="under_title">
</div>

<div id="subtitle_right">
IMS - NUS Workshop Proposal <br/>
03 | July | 2024 <br/>
Slides: <a href="https://go.wisc.edu/54tmr9">go.wisc.edu/54tmr9</a>
</div>

<div id="subtitle">
Kris Sankaran, Wei-Yin Loh <br/> 
Susan Holmes Wanjie Wang <br/>
Bibhas Chakraborty, Bee Choo Tai
</div>

---

### Outline

.pull-left[
**Setting**

1. Scientific Context
1. Statistical Challenges
1. Mathematical Opportunities
]

.pull-right[
**Structure**

1. Goals
1. Audience
1. Program
]

---

class: middle

.center[
## Scientific Context
]

---

### Data Integration

For many biological problems, it is necessary to gather complementary views,

1. Horizontal: Gather the same type of measurements across related, but not
identical cohorts.
1. Vertical: Gather multiple types of measurements from the same biological systems.

Research in data integration should make it easier to answer questions that need
data from all these sources.

---

### Example: Cancer Research

Information about the molecular activity and global organization of cells has
helped advance understanding and personalized treatment of cancer.

---

### Example: Gene Editing

(?)

---

class: middle

.center[
## Statistical Challenges
]

---

### Adaptability

How can we enable careful data analysis when each problem might involve diverse
data types and experimental designs?

1. General-purpose techniques
1. Effective diagnostics
1. Strategies for scalability

Biological insights can be hindered by inappropriate assumptions or the failure
to use all of the data.

---

### Reproducibility

Complex workflows -- which have many interlocking steps and may use black boxes
-- challenge scientific reproducibility and interpretability.

1. Assessing stability
1. Addressing selection baises
1. Supporting re-usable software

---

### Community Resources

Larger community databases can be informative, but may be noisier or only
indirectly related to questions of interest. How should they be brought into a
specific analysis?

1. Cell type atlases
1. Gene function annotation
1. Electronic health records

---

class: middle

.center[
## Mathematical Opportunities
]

---

### Motivation

Theoretical analysis has driven advances in biological data integration,

1. **Improved Algorithms: Theory can provide richer vocabulary for formulating and answering quantitative questions based on biological data.**

1. Consolidation: Theory can create coherent framework from which to compare
various available methods.

.pull-right[
  High-Dimensional CCA
]

---

### Motivation

Theoretical analysis has driven advances in biological data integration,

1. **Improved Algorithms: Theory can provide richer vocabulary for formulating and answering quantitative questions based on biological data.**

1. Consolidation: Theory can create coherent framework from which to compare various available methods.

.pull-right[
]

---

### Motivation

Theoretical analysis has driven advances in biological data integration,

1. Improved Algorithms: Theory can provide richer vocabulary for formulating and answering quantitative questions based on biological data.

1. **Consolidation: Theory can create coherent framework from which to compare various available methods.**

.pull-right[
  Tensor methods for network integration
]

---

### Motivation

Modern data can also inspire developments in theory,

1. **Expanded Abstractions: As the collection of derived applications grows, theoretical concepts can be enriched.**

1. Revised Assumptions: Properties that are taken for granted in narrower analysis settings may no longer apply.

.pull-right[
  Empirical Bayes PCA
]
---

### Motivation

Modern data can also inspire developments in theory,

1. Expanded Abstractions: As the collection of derived applications grows, theoretical concepts can be enriched.
1. **Revised Assumptions: Properties that are taken for granted in narrower analysis settings may no longer apply.**

.pull-right[
  Testing for the alignability of multiple tables
]

---

class: middle

.center[
## Workshop Structure
]

---

### Goals

---

### Audience


---


### Tutorials


---


### References

```{r, results='asis', echo = FALSE}
PrintBibliography(bib, start = 1, end = 4)
```

---

```{r, results='asis', echo = FALSE}
PrintBibliography(bib, start = 5, end = 8)
```
