title: "Marginal Analysis Example"
author: kris Sankaran
output: readthedown::html_document
---

```{r}
library(gamboostLSS)
library(scDesigner)
library(splines)
library(glue)
library(tidyverse)
```

### Using SummarizedExperiment

* How many genera are available in this experiment object?
* What was the most common phylum in this dataset?
* What was the average participant age?

```{r}
atlas <- readRDS("data/atlas.rds")

dim(exper) # 125 genera
table(rowData(atlas)$Phylum)

mean(colData(atlas)$age)

# or just,
mean(atlas$age)
```

### Gaussian Mixture

```{r}
exper_grouped <- readRDS(url("https://github.com/krisrs1128/talks/raw/master/2024/20240611/data/exper_group.rds"))
exper_histogram(exper_grouped)
```

* Explain how the simulation code below works. 
* Modify it so that it accurate reflects the difference between groups for the first three genes.

```{r}
sim <- setup_simulator(exper_grouped, ~ 1, ~ GaussianLSS()) |>
  estimate()

sample(sim) |>
  exper_histogram()
```

```{r}
sim <- sim |>
  mutate(1:3, link = ~ group) |>
  estimate()

sample(sim) |>
  exper_histogram()

```

### Gaussian over Time

```{r}
exper_ts <- readRDS("data/exper_ts.rds")
exper_lineplot(exper_ts)
```

```{r}
sim <- setup_simulator(
  exper_ts,
  ~ ns(time, df = 7) * group,
  ~ GaussianLSS()
) |>
  estimate(nu = 0.01, mstop = 1000)

sample(sim) |>
  exper_lineplot() +
```

```{r}
sim <- sim |>
  mutate(4:6, link = ~ ns(time, df = 7)) |>
  estimate(nu = 0.01, mstop = 1000)

sample(sim) |>
  exper_lineplot()
```


### Using a Simulator

```{r}
exper <- readRDS("data/atlas.rds")
sim <- setup_simulator(
  exper,
  ~ bmi_group + age + log_depth,
  ~ ZINBLSS()
) |>
  estimate(nu = 0.001, mstop = 100)
```
