---
title: "Reticulate Examples"
author: "Kris Sankaran"
date: "January 14, 2021"
output:
  xaringan::moon_reader:
    css: ["default", "css/xaringan-themer.css"]
    self_contained: false
    lib_dir: libs
    nature:
      ratio: "16:9"
    seal: false
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = T, dpi = 200, fig.align = "center", out.width = 650)
library(ggplot2)
min_theme <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    strip.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(min_theme)
```

---

class: bottom

## Reticulate Examples


.pull-left[
January 24, 2022
]
 
---

Available environments...

```{r}
library(reticulate)
conda_list()
```

---

```{r}
use_condaenv("r-reticulate")
```


---

Running python code in Rstudio.

```{python, echo = TRUE}
from sklearn import datasets, linear_model

X, y = datasets.load_diabetes(return_X_y=True)
X[:4, :5] # first five predictors
y[:4] # example response
```

---

Accessing python objects in R using `py$`.

.pull-left[
```{r}
df <- data.frame(X = py$X, y = py$y)
head(df)
```
]

.pull-right[
```{r}
library(ggplot2)
ggplot(df, aes(X.1, y)) +
  geom_point()
```

]

---

More back and forth...

.pull-left[
```{python}
model = linear_model.LinearRegression()
model.fit(X, y)
model.coef_ # fitted b coefficients
y_hat = model.predict(X)
```
]

.pull-right[
```{r}
preds <- data.frame(y = py$y, y_hat = py$y_hat)
ggplot(preds, aes(y, y_hat)) +
  geom_point()
```

]

---

We'll practice configuring a new environment,
* `conda_create` to make a new conda environment
* `conda_install` to install packages into it

You could also create / install these in the terminal and refer to them from
within R.

---

```{r}
conda_create("test")
conda_install("test", c("keras", "matplotlib"))
```

---

Here is a somewhat more involved example visualizing prediction probabilities
from a keras network.

```{python}
import matplotlib
from tensorflow import keras
from tensorflow.keras import layers
```
---

This MNIST training code is taken directly out of the keras documentation.

```{python}
import numpy as np
num_classes = 10
input_shape = (28, 28, 1)

# the data, split between train and test sets
(x_train, y_train), (x_test, y_test) = keras.datasets.mnist.load_data()

# Scale images to the [0, 1] range
x_train = x_train.astype("float32") / 255
x_test = x_test.astype("float32") / 255
# Make sure images have shape (28, 28, 1)
x_train = np.expand_dims(x_train, -1)
x_test = np.expand_dims(x_test, -1)

# convert class vectors to binary class matrices
y_train = keras.utils.to_categorical(y_train, num_classes)
y_test = keras.utils.to_categorical(y_test, num_classes)
```

---

```{python}
model = keras.Sequential([
        keras.Input(shape=input_shape),
        layers.Conv2D(32, kernel_size=(3, 3), activation="relu"),
        layers.MaxPooling2D(pool_size=(2, 2)),
        layers.Conv2D(64, kernel_size=(3, 3), activation="relu"),
        layers.MaxPooling2D(pool_size=(2, 2)),
        layers.Flatten(),
        layers.Dropout(0.5),
        layers.Dense(num_classes, activation="softmax"),
    ])
```


---

This does the training and creates the objects we'll visualize in a moment.

```{python}
model.compile(loss="categorical_crossentropy", optimizer="adam", metrics=["accuracy"])
model.fit(x_train, y_train, batch_size=128, epochs=1, validation_split=0.1)
y_hat = model.predict(x_train[:1000, :])
y = y_train[:1000, :]
```

---

```{r}
library(dplyr)
library(tidyr)
y <- data.frame(
    y_hat = py$y_hat, 
    y_true = apply(py$y, 1, function(z) which(z == 1))
  ) %>%
  mutate(i = row_number()) %>%
  pivot_longer(-i:-y_true) %>%
  mutate(
    y_true = as.factor(y_true),
    name = as.factor(name)
  )
```

---

```{r, fig.height = 5, fig.width = 6}
ggplot(y) +
  geom_histogram(aes(value, fill = name), alpha = 0.8, bins = 100, position = "identity") +
  facet_grid(y_true ~ .) +
  theme(strip.text.y = element_text(angle = 0)) +
  scale_fill_brewer(palette = "Set3")
```

---

```{r}
par(mfrow = c(2, 5))
for (i in 1:10) {
  image(py$x_train[i,,,], asp = 1)
}
```
