---
title: "Managing Batch Effects in Biological Data"
author: "Kris Sankaran"
output:
  xaringan::moon_reader:
    css: ["default", "css/xaringan-themer.css"]
    lib_dir: libs
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
    seal: false
---

class: title

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      myred: ["{\\color{myred}{#1}}", 1],
      mygreen: ["{\\color{mygreen}{#1}}", 1]
    },
    loader: {load: ['[tex]/color']},
    tex: {packages: {'[+]': ['color']}}
  }
});
</script>

<style>
.myred {color: #B4575C;}
.mygreen {color: #5A8A80;}
</style>

```{r flair_color, echo=FALSE, warning = FALSE, message = FALSE}
library(xaringancolor)
setup_colors(
  myred = "#B4575C",
  mygreen = "#5A8A80"
)

library(flair)
myred <- "#B4575C"
mygreen <- "#5A8A80"
```


```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(MASS)
library(knitr)
library(RefManageR)
library(tidyverse)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, dpi = 200, fig.align = "center", fig.width = 6, fig.height = 3)
min_theme <- theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    strip.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(min_theme)

# overwrite some default scales in ggplot2
scale_fill_continuous <- function(...) scico::scale_fill_scico(..., palette = "lapaz", direction = -1)
scale_colour_discrete <- function(...) ggplot2::scale_color_brewer(..., palette = "Set2")
scale_x_continuous <- function(...) ggplot2::scale_x_continuous(..., expand = c(0, 0))
scale_y_continuous <- function(...) ggplot2::scale_y_continuous(..., expand = c(0, 0))

BibOptions(
  check.entries = FALSE,
  bib.style = "numeric",
  cite.style = "numeric",
  style = "markdown",
  hyperlink = FALSE,
  dashed = FALSE,
  max.names = 1
)
bib <- ReadBib("references.bib")
```

## Managing Batch Effects in Biological Data

<div id="subtitle">
Kris Sankaran <br/>
(w/ Pratheepa Jeganathan)<br/>
06 | August | 2024 <br/>
Lab: <a href="https://go.wisc.edu/pgb8nl">go.wisc.edu/pgb8nl</a> <br/>
</div>

<div id="subtitle_right">
BME 780: Methods in Quantitative Biology<br/>
Slides: <a href="https://go.wisc.edu/">go.wisc.edu/</a><br/>
</div>

<!-- 45 minute talk -->

---

### Learning Outcomes


----

## Fundamental Approaches


---

## Checking your Work

---

## Discipline and Design

---

class: reference

### References

```{r, results='asis', echo = FALSE}
PrintBibliography(bib, start = 1, end = 13)
```

---

class: reference

### References

```{r, results='asis', echo = FALSE}
PrintBibliography(bib, start = 14, end = 29)
```

---



---

### Figure Attributions