---
title:  Modular software for mediation analysis of microbiome data
output:
  xaringan::moon_reader:
    css: ["default", "css/xaringan-themer.css"]
    lib_dir: libs
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
    seal: false
---

class: title

#  Modular software for mediation analysis of microbiome data

<style>
.slide-background {
    background: url("figures/cover.png") no-repeat center center;
    background-size: cover;
    opacity: 0.5;
}
</style>

<div id="subtitle_left">
Slides: <a href="https://go.wisc.edu/">go.wisc.edu/</a><br/>
Package: <a href="https://go.wisc.edu/830110">go.wisc.edu/830110</a><br/>
Lab: <a href="https://measurement-and-microbes.org">measurement-and-microbes.org</a> <br/>
</div>
<div id="subtitle_right">
Kris Sankaran <br/>
<a href="https://www.birs.ca/events/2025/5-day-workshops/25w5324/">Novel Approaches for Multi-Omics</a><br/>
17 | July | 2025 <br/>
</div>

<!-- 25 minute talk (including Q&A) -->

```{r, echo = FALSE, warning = FALSE}
library(knitr)
library(RefManageR)

opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, dpi = 200, fig.align = "center", fig.width = 6, fig.height = 3)
BibOptions(
  check.entries = FALSE,
  bib.style = "numeric",
  cite.style = "numeric",
  style = "markdown",
  hyperlink = FALSE,
  dashed = FALSE,
  max.names = 1
)
bib <- ReadBib("references.bib")
```

---

---

### Thank you!

* Contact: ksankaran@wisc.edu
* Lab Members: Margaret Thairu, Shuchen Yan, Yuliang Peng, Helena Huang
* Funding: NIGMS R01GM152744, NIAID R01AI184095

---

class: reference

### References

```{r, results='asis', echo = FALSE}
PrintBibliography(bib, start = 1, end = 13)
```
