---
title: "Methods for Biological Data Workshop [07 | February | 2024]"
author: Kris Sankaran
format:
  html:
    code-fold: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
knitr::opts_chunk$set(
  fig.align = "center",
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE, 
  eval = TRUE,
  cache = TRUE,
  dev.args = list(bg = "transparent")
)

knitr::knit_hooks$set(output = scDesigner::ansi_aware_handler)
options(crayon.enabled = TRUE)
set.seed(20240207)

library(tidyverse)
options(
  ggplot2.discrete.colour = c("#9491D9", "#3F8C61", "#F24405", "#8C2E62", "#F2B705", "#11A0D9"),
  ggplot2.discrete.fill = c("#9491D9", "#3F8C61", "#F24405", "#8C2E62", "#F2B705", "#11A0D9"),
  ggplot2.continuous.colour = function(...) scale_color_distiller(palette = "Spectral", ...),
  ggplot2.continuous.fill = function(...) scale_fill_distiller(palette = "Spectral", ...)
)

th <- theme_classic() +
  theme(
    panel.background = element_rect(fill="transparent"),
    strip.background = element_rect(fill="transparent"),
    plot.background = element_rect(fill="transparent", color=NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill="transparent"),
    legend.box.background = element_rect(fill="transparent"),
    legend.position = "bottom"
  )

theme_set(th)

heatmap <- function(...) {
  stats::heatmap(..., col=hcl.colors(15, "Zissou"))
}

suppressPackageStartupMessages(library(Matrix))
suppressPackageStartupMessages(library(SpatialExperiment))
suppressPackageStartupMessages(library(SpiecEasi))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(CovTools))
suppressPackageStartupMessages(library(gamboostLSS))
suppressPackageStartupMessages(library(genefilter))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(glue))
suppressPackageStartupMessages(library(scDesign3))
suppressPackageStartupMessages(library(splines))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(scDesigner))
```

Simulation is like visualization -- though it is usually not an end in itself,
it can deepen our thinking about a wide variety of statistical problems. In this
note, we'll specifically explore how simulation can guide microbiome analysis.
At one level, we will encounter tools for microbiome network inference and
differential hypothesis testing. At another, we'll see how to benchmarking
studies and run power analysis based on those tools, leveraging the potential
for simulation to generate realistic data.

We should distinguish between pure and semisynthetic simulators. Pure simulators
are completely specified by the designer. For example, they can include
mechanistically-informed sequence of statistical sampling and transformation
steps. These are often used in scientific and engineering domains where a small
number of mathematical equations can faithfully represent a system of interest.
In contrast, semisynthetic simulators are trained on an initial "pilot" dataset.
They learn to mimic the key properties of that source data. In some sense, they
are similar to generative models in AI, like GANs and GPTs. Unlike traditional
generative models, to be useful, these models must also be controllable
(otherwise, they would be difficult to use in downstream comparisons). Most
semisynthetic simulators in biological data analysis allow users to modify
properties of the underlying signal, dependence structure, and sample sizes.

We'll concentrate on the `scDesign` family of simulators, with an emphasis on
recent developments focused on improving interactivity. We note that 

## Longitudinal Microbiome Study

### Visualizing and Storing Data

Let's begin our study of semisynthetic using Kodikara et al.'s (2022)
longitudinal mouse microbiome dataset. The block below reads in a subset of
taxa, and the filtering script can be found at [this
link](https://github.com/krisrs1128/talks/blob/master/2024/20240207/data/longitudinal_preprocess.R).

```{r}
#| label: read
mapping <- read_csv("https://github.com/krisrs1128/talks/raw/master/2024/20240207/data/mapping.csv")
samples <- read_csv("https://github.com/krisrs1128/talks/raw/master/2024/20240207/data/filtered_samples.csv")
```

Before fitting anything, let's visualize the series. We'll have to reshape the
data into "long" format so that taxa names are stored explicitly as a variable.
The timepoints have been colored in by the phase of experiment. It seems that
most species have been affected by the antibiotic treatment and subsequent
infection, though the direction and duration of the effect can vary
substantially.

```{r}
#| label: prepare-plot
samples |>
  pivot_longer(-sample, names_to = "ASV") |>
  filter(ASV %in% glue("ASV{1:16}")) |>
  left_join(mapping) |>
  ggplot() +
  geom_line(aes(time, value, col = phase, group = subject)) +
  scale_y_log10() +
  facet_wrap(~ reorder(ASV, -value))
```

Working simply with `data.frames` can become unwieldy in this type of problem,
because there is a qualitative difference between sample-level descriptors
(e.g., sample collection time, experimental phase) and the high-throughput
sequencing outputs. This means that we should avoid combing both sources into
the same table. On the other hand, it would be nice to have a simple way of
referring to the entire experiment in a single comment. The solution we'll use
is to define a `SummarizedExperiment`. This is a special class that has slots
for the high-throughput sequencing data (`assays`) and the sample descriptors
(`colData`). Essentially, it's a container for the experiment.

```{r}
#| label: setup_experiment
exper <- SummarizedExperiment(
  assays = list(counts = t(samples[, -1])),
  colData = DataFrame(mapping)
)

rownames(colData(exper)) <- mapping$sample
colData(exper)$celltype <- rep("A", nrow(mapping)) # needed below
```

### Simulation Quick-Start

Let's begin with the `scdesign3` package. This package was originally developed
for single cell experiment studies. It turns out that its probability model is
usually able to accommodate microbiome data as well, and since the model is
generative, we'll be able to evaluate this statement precisely by comparing the
source with the simulated samples. In the call below, most of the arguments
below are placeholders. The most important argument is `mu_formula`, which says
that we should allow the mean function of the model to vary as a smoothly
varying spline over the time variable in our `colData`.

```{r}
scdesign_result <- scdesign3(
  exper,
  pseudotime = "time", 
  celltype = "celltype", 
  spatial = NULL, 
  other_covariates = NULL,
  mu_formula = "s(time, k = 10)",
  corr_formula = "1"
)
```

The simulated data are stored in the `$new_count` slot of the `scDesign3`
output. Let's visualize the simulated trajectories, keeping in mind our earlier
plot as a point of reference. We're converting the data to long format just as
before. The spline seems flexible enough to capture most of the real variation
we were seeing in these (highly abundant) taxa. We shouldn't take this for
granted! I've tried many microbiome simulators in the past, and this worked much
more straightforwardly than what's usually encountered. To get a sense of the
complexity, the sparseDOSSA2 paper of Ma et al. (2023) recommends fitting a
separate simulator to every timepoint in a longitudinal dataset.

```{r}
data.frame(t(scdesign_result$new_count)) |>
  rownames_to_column("sample") |>
  pivot_longer(-sample, names_to = "ASV") |>
  filter(ASV %in% glue("ASV{1:16}")) |>
  left_join(mapping) |>
  ggplot() +
  geom_line(aes(time, value, col = phase, group = subject)) +
  scale_y_log10() +
  facet_wrap(~ reorder(ASV, -value))
```

### `scDesigner`

`scDesign3` is well-documented and gives good results in many microbiome
problems. It's a good default for any current simulation-based analysis.
Nonetheless, we will introduce some of the main ideas of `scDesigner`, a
follow-up project that's under active development. The main idea of `scDesigner`
is to create a vocabulary for interactively building, using, and critiquing
models from the `scDesign3` model. The `scDesign3` copula model is actually a
very interpretable model, it would be a shame to treat the `scdesign3` function
that we called above as if it were a black box.

`scDesigner` supports the following operations:

* Creation: We can use a formula interface to specify the relationship between experimental covariates and LSS/Copula model parameters.
* Alteration: Given an existing simulator, we can modify the regression formula, probability family, and dependence structure, and re-estimate just those parts that have changed.
* Sampling: We can sample synthetic `SummarizedExperiment` classes from the fitted model. 
* Joining: Given a pair of previously defined simulators, we can unify them into a single simulator. This is especially helpful in multi-omics settings, because it lets us first refine our simulators separately on each omic before finally modeling their joint dependence structure.
* Plotting: We can use visualization to compare original vs. simulated data, which is useful for refining and improving simulators.

Let's see this in action. We'll first define the model for each taxon. We're
using a negative binomial location-shape-scale model where the parameter can
vary with experimental `phase`. This is more constrained than the full temporal
model we defined above. The intent is to see to what extent `phase` is enough to
describe the essential variation in the data (we should always prefer a more
parsimonius explanation if possible).

```{r}
margins <- setup_margins(exper, ~ phase, ~ NBinomialLSS())
margins
```

`setup_margins` species the regression relationships without actually fitting
the model. We can visualize the fitted regression together with the original
training samples.

```{r}
fit <- estimate(margins, exper)
plot(fit, exper)
```

We can generate a synthetic dataset using `sample`, and we can directly contrast
them with the original data by using the `simulated` argument to `plot`. It
seems like there are important variations over time that are not being captured
by `phase` alone. For example, ASV8 has a decrease during the infection period,
but our current model only allows for a fixed value over those last few days.

```{r}
#| fig-width: 10
#| fig-height: 4
fit <- estimate(margins, exper)
plot(fit, exper, simulated = sample(fit))
```

To address this let's allow the mean to vary as a natural spline over time.
We'll fix the variance over all timepoints. This is helps to stablize the
estimated model, at the expense of less expressive fits. Overall, we're much
better now at capturing trends, but still struggle in some of the very large
positive values (ASV2, ASV3) and also zero-inflation (ASV1, ASV7, ASV8). It
would be possible to improve the fit by switching from a `NBinomialLSS()` model
to `ZINBLSS()`. This model takes a bit longer to fit, so we'll skip it in this
note, though you're of course encouraged to try exploring the change yourself.

```{r}
#| fig-width: 10
#| fig-height: 4
fit <- margins |>
  mutate(link = list(mu = ~ ns(time, 10), sigma = ~ 1)) |>
  estimate(exper)
plot(fit, exper, simulated = sample(fit))
```

Let's visualize the full trajectories for all taxa. Here, we're using the
`pivot_experiment()` function, which converts a `SummarizedExperiment` class
into a long `data.frame` for easier plotting. In many ways, these samples
convincingly capture many of the salient properties of the original data. The
deficiencies are similar to what we noted before -- we have trouble capturing
zeros and changes in variance. In lower-stakes applications, this might not be
too much of a problem, and these data could still guide choices around
methodology or experimental design. In higher-stakes settings, we could modify
the probability model and regression family using the `mutate` function until we
arrive at a more satisfying simulator. We emphasize again that it's the ability
to simulate that's making it so easy to critique these models, because comparing
real with simulated samples is a much simpler task than inspecting model
parameters directly.

```{r}
#| fig-width: 10
#| fig-height: 6
sample(fit) |>
  pivot_experiment() |>
  ggplot() +
  geom_line(aes(time, log(1 + value), col = phase, group = subject), alpha = 0.4) +
  guides(col = guide_legend(override.aes = list(linewidth = 2, alpha = 1))) +
  facet_wrap(~ reorder(feature, -value), ncol = 10)
```

## Microbiome Networks

Unlike human social networks, there is no simple way to observe microbe-microbe
interactions -- we have to make do with indirect evidence. One approach uses
population profiles as a proxy for ecological interaction. Taxa that often
co-occur are understood to have cooperative ecological interactions, while those
that don't are thought to compete for the same niche.

Many algorithms have been designed around this intuition, all trying to go
beyond simple co-occurrence and instead capture more complex types of
dependence. A challenge in practice is that it's hard to know which method to
use when, since the problem is unsupervised. Even when thorough simulation
benchmarking studies are available, it's often not obvious how well those
simulation setups match our problems of interest.

Let's use `scDesigner` to benchmark network estimation methods using data from
rounds 1 and 2 of the [American Gut
Project](https://journals.asm.org/doi/10.1128/msystems.00031-18). We will
simulate data with known correlation structure and taxa-level marginals
estimated from the study data. The block below reads in the data.

```{r}
#| label: read-data
counts <- read_csv("https://github.com/krisrs1128/talks/raw/master/2023/20231206/exercises/amgut-counts.csv") |>
  column_to_rownames("taxon")
metadata <- read_csv("https://github.com/krisrs1128/talks/raw/master/2023/20231206/exercises/amgut-metadata.csv")

exper <- SummarizedExperiment(
  assays = SimpleList(counts = as.matrix(counts)),
  colData = metadata
)
```

First, let's estimate a simulator with the taxon-wise regression formula
`~log(sequencing_depth) + BMI`. We will use a zero-inflated negative binomial
location-shape-scale (`ZINBLSS`) models for each gene, using a gaussian copula
to capture dependence. The data structure below captures all the simulator
components, and we can swap pieces in and out to modify the form of the
simulator. For example, if we wanted, we could `mutate` the family and link
function associated with particular features.

```{r}
#| label: example-sim
sim <- setup_simulator(
  exper,
  ~ log(sequencing_depth) + BMI,
  ~ ZINBLSS()
) |>
  estimate(mstop = 100)
sim
```

The simulated data is always a `SummarizedExperiment`. This means that any
workflow that applied to the original data can be applied to the simulated one
without any changes. Notice also that `sample` defaults to drawing samples from
the same design as the original input experiment (we'll modify this using the
`new_data` argument in a minute).

```{r}
simulated <- sample(sim)
simulated
```

Let's compare the marginal count distributions for the real and simulated data.
We'll need the data in "long" format to be able to make the ggplot2 figure. The
`pivot_experiment` helper can transform the original `SummarizedExperiment`
objects in this way. Notice that the simulated data tends to overestimate the
number of zeros in the high-abundance taxa. To refine the simulator, we should
probably replace the zero-inflated negative binomial with ordinary negative
binomials for these poorly fitted taxa.

```{r}
#| label: histogram-critique
#| fig-width: 9
#| fig-height: 5
bind_rows(
  real = pivot_experiment(exper),
  simulated = pivot_experiment(simulated),
  .id = "source"
) |>
  filter(feature %in% rownames(simulated)[1:20]) |>
  ggplot() +
  geom_histogram(
    aes(log(1 + value), fill = source),
    position = "identity", alpha = 0.8
  ) +
  facet_wrap(~ reorder(feature, value), scales = "free")
```

Are the learned relationships with BMI plausible? We can compare scatterplots of
the real and simulated data against this variable. Note that, by default, the
ribbons will be evaluated along all variables, which makes for the jagged
ribbons (neighboring values for BMI might have different sequencing depth,
potentially leading to quite different predictions). To remove this artifact, we
can assume that all samples had exactly the same sequencing depth.

```{r}
#| label: bmi-scatter
#| fig-width: 12
#| fig-height: 4
new_data <- colData(exper) |>
  as_tibble() |>
  mutate(sequencing_depth = 2e4)
plot(sim, "BMI", sample(sim, new_data = new_data), new_data = new_data)
```

Next, let's visualize the correlation matrix estimated by the simulator's copula model.

There are a few pairs of taxa that are very highly correlated, and there are
also a few taxa that seem to have higher correlation across a large number of
taxa (e.g., the taxon in row 34). There is no obvious banding or block structure
in this real data, though.

```{r}
#| label: correlation-heatmap
#| fig-width: 5
#| fig-heigth: 5
rho <- copula_parameters(sim)
heatmap(rho)
```

Next, let's pick a pair of taxa with high correlation under this model and
visualize the joint distribution of their ranks to see if they are plausible.
I've picked one of the higher-correlation pairs, and you can replace the
selection with the commented out line to see what one of the anticorrelated
pairs of taxa looks like.

```{r}
#| label: correlation-pairs
#| fig-width: 5
#| fig-heigth: 5
#taxa <- rownames(exper)[c(33, 43)]
taxa <- rownames(exper)[c(14, 25)]
pivot_experiment(exper) |>
  filter(feature %in% taxa) |>
  pivot_wider(names_from = feature) |>
  ggplot() +
  geom_point(aes(log(1 + .data[[taxa[1]]]), log(1 + .data[[taxa[2]]])))
```

Finally, let's define a ground truth correlation structure that can be the basis
of our benchmarking experiment. We'll replace the current copula correlation
structure with one from a block diagonal matrix. In this example, the
off-diagonal correlations are 0.6. We can use `mutate_correlation` to swap this
new correlation matrix into our earlier simulator.

```{r}
#| label: mutate-correlation
rho <- c(0.4, .6, 0.8) |>
  map(~ matrix(., nrow = 15, ncol = 15)) |>
  bdiag() |>
  as.matrix()
diag(rho) <- 1

simulated <- sim |>
  mutate_correlation(rho) |>
  sample()

x <- t(assay(simulated))
```

Let's first look at the [`SpiecEasi` covariance estimate](https://github.com/zdk123/SpiecEasi). This is a variant of the
graphical lasso that is designed to be well-adapted to microbiome data. The good
news is that it does warn that the default choices of $\lambda$ are too large,
which is correct in this case. Unfortunately, it took a while to get this
answer, and we had already been quite generous in allowing it to fit 10 choices
of $\lambda$.

```{r}
#| label: spieceasi
rho_se <- spiec.easi(x, nlambda = 10, pulsar.params = list(rep.num = 1)) |>  getOptCov() |>
  as.matrix() |>
  cov2cor()
heatmap(rho_se)
```

Let's instead use the Ledoit-Wolfe estimator on the log-transformed data. The
results make much more sense.

```{r}
#| label: lw
rho_lw <- CovEst.2003LW(log(1 + x))$S |>
  cov2cor()
heatmap(rho_lw)
```

Since color comparisons are difficult to evaluate precisely, we can also make a
scatterplot comparing the different covariance estimators.

```{r}
#| label: covariance-metrics
data.frame(truth = c(rho), se = c(rho_se), lw = c(rho_lw)) |>
  pivot_longer(-truth, values_to = "estimate") |>
  ggplot() +
  geom_jitter(aes(truth, estimate, col = name), alpha = 0.6, size = 0.4) +
  geom_abline(slope = 1) +
  facet_wrap(~name)
```

This example shows that, when we start with real template data, it's not too
hard to setup a benchmarking experiment. It's generally easier to reconfigure
the components of an existing simulator than it is to specify all the simulation
steps from scratch. There is the secondary bonus that the data tend to look
close to real data of interest, at least up to the deliberate transformations
needed to establish ground truth.

We could imagine extending this example to include different data properties
(sample sizes, variable block sizes and correlations, more general correlation
structure) and estimation strategies (alternative transformations or
estimators). Design changes could be implemented using `expand_colData`, changes
in the signal can be specified as above with `mutate_correlation`, and any
workflow can be used as long as it applies to a `SummarizedExperiment` object.

### Differential Testing Power Analysis

In microbiome studies, the goal of differential testing is to look
taxon-by-taxon (or, in metagenomics data, gene-by-gene) for associations with
some sample-level characteristic of interest. We deliberately ignore dependence
across taxa to clarify which marginal relationships are worth attending to more
carefully. One advantage of this perspective is that it matches nicely with
experimental capabilities: it's easier to manipulate individual taxa/genes in
follow-up work than attempting to change the entire system.

A common problem when first embarking on a difeferential testing study is to run
a power analysis. A power analysis is about more than choosing the sample size.
First, it can help guide allocation of a fixed sampling budget. For example, we
might realize that a design that pairs pre-post samples might be more powerful
than a cross-sectional one with the same total number of samples. Second, it can
inform the testing approach -- power is impossible to disentangle from analysis
strategy. A simulation experiment might suggest that a simpler analysis will
actually have higher power, given sampling budget constraints.

To explore the main ideas of power analysis, let's imagine that we are planning
a new study to evaluate the relationship between the gut microbiome and colon
cancer. Simulating from scratch is challenging, so we refer to the
curatedMetagenomicsDatasets online database to download preprocessed data from
Yachida et al. (2019). This will be the pilot data for our semisynthetic
simulator. The download and filtering script can be found
[here](https://github.com/krisrs1128/talks/blob/master/2024/20240207/data/yachida_preprocess.R).

The block below encapsulates differential testing using `limma-voom`. This is a
classic approach to differential testing in RNA-seq that has been found to be
reasonably effective in microbiome studies (though, there are many
microbiome-specific competitors to keep in mind these days, I recommend the
benchmarks by Calgaro et al. 2020 and Cappellato et al. 2022). It's built on the
ideas that (i) samples should be normalized to approximately the same scale
(making an effort to avoid compositional biases), and (ii) considering
feature-level mean-variance relationships can make classic linear modeling
testing tools applicable, so that we can bypass fancier -- and often more
finnicky -- count-based probabilistic models.

```{r}
differential_test <- function(dge, col_data, max_p = 1) {
  design <- model.matrix(~ disease, data = col_data)
  voom(dge, design) |> # account for differences in uncertainty
    lmFit(design) |>
    eBayes() |>
    topTable(p.value = max_p, number = Inf) |>
    arrange(adj.P.Val) |>
    rownames_to_column("ID")
}
```

Let's run that differential testing pipeline and create a volcano plot of fold
change vs. statistical significance. A few taxa seem clearly different group
groups. The question now is how we can use the results from this pilot
experiment to define a power analysis.

```{r}
yachida <- readRDS(url("https://github.com/krisrs1128/talks/raw/master/2024/20240207/data/yachida.rds"))
dge <- DGEList(assay(yachida), group = colData(yachida)$disease) |>
  calcNormFactors()

results <- differential_test(dge, colData(yachida))
ggplot(results, aes(logFC, -log(adj.P.Val, 10))) +
  geom_point() +
  geom_text_repel(data = filter(results,  abs(logFC) > 2), aes(label = ID), size = 3)
```

To evaluate power, we need a sense in which some taxa are true signals/true
nulls. We don't know the ground truth in the real data, but we can create ground
truth in our simulation. Moreover, since our simulators are synthetic, we can
ensure that the generated samples are plausible. To train the simulator, we'll
work with the `limma-voom` normalized samples. This is the role of the call to
`cpm`, which stands for "Counts-per-Million" and implements the normalization.

```{r}
assays(yachida, withDimnames = FALSE)$normalized <- cpm(dge, log = TRUE)
sim <- setup_simulator(yachida,  ~ disease, ~ GaussianLSS()) |>
  estimate(yachida, "normalized")
```

Next, we'll create ground truth nulls. We'll use the 60 taxa that already seemed
to have only weak effects in the pilot experiment. By making the link function
`~ 1` instead of `~ disease`, we ensure that there is in fact no association
between disease and abundance for those taxa.

```{r}
nulls <- tail(results$ID, 60)
sim <- sim |>
  mutate(any_of(nulls), link = ~ 1) |>
  estimate(yachida, "normalized")
```

From here, we can sample from the simulator and rerun our differential testing
function. We can modify the sample size $N$ and see how power and the fraction
of false positives vary. If we ran more replicates and choices of $N$, we would
be able to draw a smooth curve of how these metrics vary as a function of $N$,
like what we would expect theoretically. The advantage of this computational
setup is that calculating this curve analytically would be impossible, given the
complexity of microbiome data, and since our simulator has been trained to mimic
a pilot dataset, we should be able generate more realistic samples than if we
had defined the sampling mechanism from scratch.

```{r}
N <- 500
assignments <- c(rep("CRC", N / 2), rep("healthy", N / 2))
new_data <- tibble(disease = factor(assignments, levels = c("healthy", "CRC")))
simulated <- sample(sim, new_data = new_data)

new_results <- DGEList(2 ^ assay(simulated), group = colData(simulated)$disease) |>
  differential_test(colData(simulated)) |>
  mutate(null = ID %in% nulls)

new_results |>
  ggplot(aes(logFC, -log(adj.P.Val, 10), col = null)) +
  geom_point() +
  geom_hline(yintercept = -log(0.05, 10), col = "#ababab") +
  geom_text_repel(data = filter(new_results,  abs(logFC) > 1.5), aes(label = ID), size = 3) +
  facet_wrap(~ null)
```

### Augmenting Spatial Data

Modern molecular biology aims to build a map of the cell, its molecular
machinery, and its relationships within larger systems (e.g., the circulatory
system). Yet, until 2018 - 2020, this map had no sense of spatial orientation.
This has changed with the advent of spatial sequencing technologies, where we
can now measure the abundances of individual molecules together with their
spatial location within tissue samples.

In this exercise, we will build a model of spatial transcription in the
dorsolateral prefrontal cortex (DLPFC). This model could then be used downstream
to create negative controls and ensure properly calibrated spatial differential
expression analysis. The block below reads a small subset of [DLPFC
data](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8095368/) into an object of
class [`SpatialExperiment`](https://github.com/drighelli/SpatialExperiment). 

```{r}
#| label: spatial-estimation
spatial_experiment <- readRDS(url("https://github.com/krisrs1128/talks/raw/master/2023/20231206/exercises/dlpfc-v2.rds"))
```

First, let's visualize the spatial gene expression pattern associated with a subset of
genes. We would like to understand the extent of spatial smoothness and skewness
in expression levels. For the genes that we've selected, there is a trend where
the expression levels increase along individual rows but don't change too much
along a single column, once they're within a certain closed region (though see
the refinements below).

```{r}
#| label: raw-data
coords <- spatialCoords(spatial_experiment)

t(assay(spatial_experiment)[1:3, ]) |>
  as.matrix() |>
  bind_cols(coords) |>
  pivot_longer(-starts_with("pxl")) |>
  ggplot() +
  geom_point(aes(pxl_col_in_fullres, pxl_row_in_fullres, col = log(1 + value))) +
  facet_wrap(~ name)
```

Next, let's estimate a simulator based on these observations. The simulator
formulas can only refer to columns in the experiment's `colData`. This makes
this problem a bit tricky, since the relevant spatial predictors are not
available in this way.

```{r}
#| label: original-names
colnames(colData(spatial_experiment))
```

We could always manually input them using the same `spatialCoords` function that
we used to make the plot above. There is a more direct way using `augment`, made
possible because `scDesigner` can check experiment data classes in order to
identify relevant variables for modeling:

```{r}
#| label: augmented-names
spatial_experiment <- augment(spatial_experiment)
colnames(colData(spatial_experiment))
```

We'll just use the interaction between two univariate spline functions. We're
using negative binomial models -- it's educational to try out other families
(e.g., `GaussianLSS`, `StudentTLSS`, or `ZIPoLSS`).

```{r}
#| label: fit-spatial
simulator <- spatial_experiment |>
  setup_margins(~ ns(pxl_col_in_fullres, 3) * ns(pxl_row_in_fullres, 3), ~ NBinomialLSS()) |>
  estimate(spatial_experiment, mstop = 50)
```

We can visualize the results using `plot_bivariate`. The smoothed values reveal
that there is a subtle increasing trend from lower to higher
`pxl_row_in_fullres` values within the main expression blocks -- this was hard
to tease out from the original data and is one of the benefits of fitting a
model.

```{r}
#| label: vis-spatial
#| fig-height: 7
#| fig-width: 9
features <- c("pxl_col_in_fullres", "pxl_row_in_fullres")
plot_bivariate(simulator, spatial_experiment, features, max_print = 20) +
  labs(
    "fill" = expression(log(1 + count)),
    "color" = expression(log(1 + count))
  )
```

Given this simulator, we can imagine planting negative controls by removing the
spatial structure for some genes. This can be helpful for calibrating spatial
differential expression tests. Alternatively, we could evaluate spatial
clustering methods by defining a ground truth where many genes are simulated
from exactly the same probability model.

## References

Calgaro, Matteo, Chiara Romualdi, Levi Waldron, Davide Risso and Nicola Vitulo. Assessment of statistical methods from single cell, bulk RNA-seq, and metagenomics applied to microbiome data. Genome Biology 21 (2020): n. pag.

Cappellato, Marco, Giacomo Baruzzo and Barbara Di Camillo. Investigating differential abundance methods in microbiome data: A benchmark study. PLoS Computational Biology 18 (2022): n. pag.

Kodikara, Saritha, Susan Ellul, Kim-Anh Lê Cao, Statistical challenges in longitudinal microbiome data analysis, Briefings in Bioinformatics, Volume 23, Issue 4, July 2022, bbac273, https://doi.org/10.1093/bib/bbac273

Ma, Siyuan, Boyu Ren, Himel Mallick, Yo Sup Moon, Emma H. Schwager, Sagun Maharjan, Timothy L. Tickle, Yiren Lu, Rachel N. Carmody, Eric A. Franzosa, Lucas Janson and Curtis Huttenhower. A statistical model for describing and simulating microbial community profiles. PLoS Computational Biology 17, 2021. https://doi.org/10.1371/journal.pcbi.1008913
