---
title: "Interpretability: Motivation and Definitions"
author: "Kris Sankaran"
output:
  xaringan::moon_reader:
    css: ["default", "css/xaringan-themer.css"]
    lib_dir: libs
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
    seal: false
# output: html_document
---

class: title

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      reals: "{\\mathbb{R}}",
      indic: ["{\\mathbf{1}\\left\\{{#1}\\right\\}}", 1],
      Esubarg: ["{\\mathbf{E}_{#1}\\left[{#2}\\right]}", 2],
      absarg: "{\\left|{#1}\\right|}",
      "\*": ["{\\mathbf{#1}}", 1],
      diag: ["{\\text{diag}\\left({#1}\\right)}", 1]
    },
    loader: {load: ['[tex]/color']},
    tex: {packages: {'[+]': ['color']}}
  }
});
</script>


```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(MASS)
library(knitr)
library(RefManageR)
library(tidyverse)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, dpi = 200, fig.align = "center", fig.width = 6, fig.height = 3)

BibOptions(
  check.entries = FALSE,
  bib.style = "numeric",
  cite.style = "numeric",
  style = "markdown",
  hyperlink = FALSE,
  dashed = FALSE,
  max.names = 1
)
bib <- ReadBib("references.bib")
```

## Visualizing Microbiome Data

<div id="subtitle">
Kris Sankaran <br/>
15 | January | 2025 <br/>
Lab: <a href="https://go.wisc.edu/pgb8nl">go.wisc.edu/pgb8nl</a> <br/>
</div>

<div id="subtitle_right">
<br/>
Slides: <a href="https://go.wisc.edu/">go.wisc.edu/</a><br/>
</div>
<!-- 30 minute talk -->

---

"Premature summarization is the root of all evil in statistics"

---

## Principles

---

### Information Density

---

### Focus-plus-Context

Also called iterative structuration.

---

### Coordinated Views

---

## Applications

---

### Differential Analysis

---

### Solution: MA Plots

---

### Solution: Volcano Plots

---

### Community Dynamics

Problem: 
Visual Solution(s):
How to read it:
Limitations:
Code example: 

---

### Solution: Principal Components Analysis

---

### Solution: Topic Analysis

---

### Solution: Transfer Analysis

---

### Multiscale Analysis

---

### Solution: Topic Alignment

---

### Data Integration

---

### Solution: Canonical Correlation Analysis

---

### Solution: Mediation Analysis

---

### Batch Effect Detection

---

### Solution: Percentile Plot

---

### Experimental Design

---

### Solution: Power Curves

---

class: reference

### References

```{r, results='asis', echo = FALSE}
PrintBibliography(bib, start = 1, end = 13)
```
